---
title: "Midpoint Deliverable"
author: "Daniel Chen, Grace Teebken, Tanner Lauzon, Kellen Maier"
date: "05/26/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# import natural disaster and average temp by country cleaning functions
source("scripts/data_cleaning.R")
# import summary info calculation functions
source("scripts/summary_info.R")
# import aggregate table calculation functions
source("scripts/aggregate_table.R")
# import plot 1
source("scripts/plot1.R")
# import plot 2
source("scripts/plot2.R")
# import plot 3
source("scripts/plot3.R")

# load libraries
library(dplyr)
library(ggplot2)
library(countrycode)
library(knitr)
```

## Introduction
This project will focus on global climate change impacts and consider temperature changes as well as natural disaster events. We aim to visualize trends in global disasters and temperature data. This will help illuminate areas that have been impacted by climate change the most. We hope that this project will lead to greater awareness of the extent of climate change impacts as well as an understanding of which areas need the most attention and resources for combating the effects of climate change.

## Data Cleaning
We will be examining **two datasets** for this deliverable. The first consists of *yearly average land temperatures per country*. The other consists of a combination of *yearly global average land temperatures* and *yearly count and total damages of natural disaster events*. We have already gathered the necessary data cleaning functions in `{r setup}` from `scripts/data_cleaning.R` and below we first gather all of the seperate csv files and load them in as dataframes.
```{r, cache=TRUE}
# DATASET 1
# gather average land temperature by country
avg_temp_by_country <-
  read.csv("data/kaggle_global_temp/GlobalLandTemperaturesByCountry.csv",
           stringsAsFactors = FALSE)

# DATASET 2
# gather global temperature averages
global_temp <-
  read.csv("data/kaggle_global_temp/GlobalTemperatures.csv",
           stringsAsFactors = FALSE)

# gather natural disaster counts
natural_disaster_count <-
  read.csv("data/kaggle_natural_disaster/number-of-natural-disaster-events.csv",
           stringsAsFactors = FALSE)

# gather natural disaster USD damages
natural_disaster_damage <-
  read.csv("data/kaggle_natural_disaster/damage-from-natural-disasters.csv",
           stringsAsFactors = FALSE)
```


We then clean up and combine the individual dataframes to get our final datasets that we will be using in our further analyses.
```{r, cache=TRUE}
# DATASET 1
# get clean average temperature by country dataset
avg_country_temp_data <-
  get_clean_avg_country_temp(avg_temp_by_country)

# DATASET 2
# get clean combined natural disaster count, damage, and temp dataset
natural_disaster_data <-
  get_clean_natural_disaster(global_temp,
                             natural_disaster_count,
                             natural_disaster_damage)
```


## Summary Information

We can now gather some brief summary information about our two datasets. First, we look at our dataset 1, the average country temperature dataset, and look into the number of observations, its breadth in terms of coverage by years and regions, and some key facts about the dataset.

```{r}
# gather summary info for DATASET 1 - average country temperature
temp_summary_info <- get_summary_info_temp(avg_country_temp_data)
temp_summary_info
```

Our first dataset, which has average land temperatures by country, has **`r format(temp_summary_info$length, big.mark = ",")` measurements** from **`r temp_summary_info$range$min_year` to `r temp_summary_info$range$max_year`**. It provides data for **`r temp_summary_info$n_regions` countries and territories**. The country with the highest average land temperature for all years is **`r format(temp_summary_info$country_highest_all_years$region)`** with an average land temperature of **`r round(temp_summary_info$country_highest_all_years$avg_temp, 1)` ˚C** as measured for the monthly average land temperature of the `r format(temp_summary_info$country_highest_all_years$month)`th month of `r format(temp_summary_info$country_highest_all_years$year)`.

In 1988, Dr. James Hansen [testified to Congress](https://www.nytimes.com/1988/06/24/us/global-warming-has-begun-expert-tells-senate.html) that global warming was "99 percent certain" and that the world needed to take action to prevent climate disaster. His testimony put a spotlight on the topic of climate change, and since then, climate change has been an important issue. This important year marks the point at which the American public became aware of the dangers of climate change and realized that action needed to be taken.

As such, we find it important to examine this dataset via two angles, before or during 1988 and after 1988, and identify the countries with the highest average land temperatures to examine potential impacts of his speech. Considering years before and including 1988, the country with the highest average temperature was **`r format(temp_summary_info$country_highest_before_1988$region)`** with an average land  temperature of **`r round(temp_summary_info$country_highest_before_1988$avg_temp, 1)` ˚C** as measured for the monthly average land temperature of the `r format(temp_summary_info$country_highest_before_1988$month)`th month of `r format(temp_summary_info$country_highest_before_1988$year)`. Considering years after 1988, the country with the highest average temperature was **`r format(temp_summary_info$country_highest_after_1988$region)`** with an average land temperature of **`r round(temp_summary_info$country_highest_after_1988$avg_temp, 1)` ˚C** as measured for the monthly average land temperature of the `r format(temp_summary_info$country_highest_after_1988$month)`th month of `r format(temp_summary_info$country_highest_after_1988$year)`. Interestingly, Kuwait is the country that holds the highest monthly average land temperature for both before or during 1988 and after 1988 even though it does not exactly land on the equator. Unfortunately, it appears that Dr. James Hansen's speech did not halt or reverse the impact of climate change. Furthermore, while considering years before or during 1988 the highest temperature was recorded in 1921. If we were expecting temperatures to continously increase over time we should expect the highest temperatures in the before or during 1988 time range to occur near 1988, yet it occurs in 1921 more than 60 years before our cutoff. Perhaps there was a fluke natural event that caused the monthly temperature to shoot up, but this is all just speculation and we cannot be too sure of any hypothesis without further testing. However, for the post 1988 time range, our highest monthly mean land temperature recording was in 2012 which is fairly close to the cutoff of 2013, our latest year in that time range, which does correspond with the idea that a continous increase in temperature over time should place the highest temperature measurment as late as possible. Of note, is the month of all of these measurments, all three (all years, before or during 1988, and after 1988) of these time ranges have their highest measurments in July, the 7th month, as such it may be interesting to group by the year and average the temperature across months to mitigate monthly variation and examine if the spike in July temperature also indicates with an increase in the yearly temperature.

--------

We can then gather some summary information, shown below, from the second dataset, count and damage caused by natural disasters along with their associated global land temperatures. Similarly to the previous summary information we look at the *number of observations* in this dataset along with its breadth in terms of *coverage by years*. Furthermore, we find the years with the *highest economic damage* by natural disasters, *highest number of natural disasters*, *highest natural disaster impact* (defined as damage / impact for each naturla disaster event) and years with the *highest associated global mean land temperature*. To do this, we only look only at the rows with *All natural disasters* as the `disaster` type in order to give a broad overview of all natural disasters without having to compare between all of the individual natural disasters.

```{r}
# gather summary info for DATASET 2 - natural disaster count, damage,
#   and associated global temperature
disaster_summary_info <- get_summary_info_disaster(natural_disaster_data)
disaster_summary_info
```

From the above, we can see that our second dataset has `r disaster_summary_info$length` descriptions of yearly disasters. Its breadth in terms of a range of years is `r disaster_summary_info$range$min_year` to `r disaster_summary_info$range$max_year`. The year with the highest global mean land and ocean temperature was `r disaster_summary_info$year_highest_temp$year` with a temperature of `r round( disaster_summary_info$year_highest_temp$mean_land_ocean_temp, 1)` degrees Celsius. In this case we are using the mean land and ocean temperature as our measurment for the global temperature because it takes a more holisitic view of the global condition rather than just focusing on the mean land temperature. While 2015 may have just been a fluke, this does to seem to point to an increase in global mean temperature over time as the most recent year in the dataset is also the hottest. The year with the most natural disasters was `r disaster_summary_info$year_highest_count$year`, with a total of `r disaster_summary_info$year_highest_count$count` disasters. The year with the most economic damage due to natural disasters was `r disaster_summary_info$year_highest_economic_damage$year`. The most impacted year (most economic damage per disaster) was `r disaster_summary_info$year_highest_impact$year`, with an average cost of $`r sprintf("%.2f billion", disaster_summary_info$year_highest_impact$impact / 1000000000)` per disaster. Interestingly, these three different calculations, though describing similar data and measurments, give three different years. It would be of value to examine what makes these three years so different, and to see if there are any similar characteristics that tie them together to reveal why one has high counts, the other high economic damage, and the last high damage per natural disaster.


## Aggregate Table
Next, we look at our dataset 1, the average land temperature, via an aggregate table, shown below, to examine how temperatures differed and changed across time. To examine this dataset we consider three chunks of time, an **ancient** time period from *1842 to 1872*, a **past** time period from *1912 to 1942*, and a **present** time period from *1982 to 2012*. We will examine the average temperatures in ˚C for each of these time period and how temperature has *changed between the past and ancient, and the present and past time periods*. Then we can examine the *overall change in temperature*, of which we will sort this dataframe by to look at countries that were most impact (i.e. changed the most across time). Furthermore we compute the *difference between the change in temperature from the past to present time periods and the change in temperature from the ancient to past time periods*. This allows us to see if, recently, temperature have been changing faster than normal.

```{r}
# gather aggregate table for DATASET 1 - average country temperature
temp_aggr_table <- get_aggr_table_temp(avg_country_temp_data)
kable(temp_aggr_table)
```

When looking at the table above we see that the *East Asia & Pacific* region of the world has the greatest overall change in mean land temperature. However, they have appeared to slow down the rate at which their average temperature increases, as seen in the `Changes Between Recent and Previous Changes (˚C)` column, where they have a value of -1.858 ˚C, meaning that their most recent change (past to present) in temperature is 1.858 ˚C smaller than their previous change (ancient to past) in temperature. In addition, the *Middle East & North Africa* region has the second highest overall change in mean land temperature and also has a negative change between their previous and recent change in temperatures. It would be interesting to see if there are reasons why these two regions have decreased their rate of increase in temperature across time, perhaps the countries in the respective areas are pushing hard on environmental fronts but we cannot be sure until further research is done. However, even though some of these regions are decreasing in their rate of temperature increase across time they all have positive overall changes in mean land temperature meaning all of these regions are increasing their mean land temperature across time. Of other interest is that, while the *Latin America & Caribbean* region has the lowest overall change in mean land temperature it has the highest change between its recent and previous change in temperatures. This tells us that it has recently shot up in the rate at which it increases its mean land temperature, and it would be interesting to understand why as its previous change from the ancient and previous time periods was negative.

--------

We then examine our dataset 2, the combined natural disaster dataset via an aggregate table, shown below, to view how natural disasters have *changed* both in their number of *occurences*, their total *damage*, and how *global temperatures* have also changed. Using an average of **30 years in the past**, *1912 to 1942*, and an average of **30 years in the present**, *1982 to 2012*, we compute the change in count and damage of natural disasters, and the associated change in global mean land and land+ocean temperatures. Furthermore, we computed the *impact* of each natural disasters by dividing its damage for the year by the number of occurences. This provides us with a quantitative measurment for the amount of damage the natural disaster causes per incident, and allows us to examine if a disaster is causing more damage per incident as time goes on. This aggregate table is then sorted from highest to lowest, descending order, by the *difference in economic damage* to focus on which types of natural disasters cause the greatest yearly damage on average.

```{r}
# gather aggregate table for DATASET 2 - natural disaster count, damage,
#   and associated global temperature
disaster_aggr_table <- get_aggr_table_disaster(natural_disaster_data)
kable(disaster_aggr_table)
```

As shown in the table above, we have also included the **All natural disasters** row as it gives a easier summary of how the total count and damage of natural disasters have changed throughought time along with their associated change in global mean temperatures. We can see from this **All natural disasters** row that not only has the global temperature increased, but the number of natural disasters along with their total yearly damage and impact (damage per incident) have also increased as well. While, **extreme weather** is the natural disaster with the *greatest increase in economic damage* (increased by ~28.8 billions of USD), earthquakes have the highest change in impact. This means that in the most recent 30 years, 1982 to 2012, **earthquakes** are now causing *significantly more damage per event* (increased by ~800 million per incident) than the past group of 30 years, 1912 to 1942. Of further concern is the universal increase in count of natural disasters, while **floods** have actually *decreased in impact*, perhaps we are now better equipped to mitigate its damage, they have *severely increased in the number of average occurences per year*. Overall, we can see that natural disasters, along with global temperatures, have increased both in their count and damage across time.


## Data Visualizations
This first chart shows the relationship between location and change in temperature from 1912 to 2012. The color coded map shows that over the last 100  years many countries have seen a rise in temperature. Specifically the northern half of the map has had the biggest noticeable rise and as showin below in the "Disaster Count by Mean Land Temp" plot, rise in temperature is unwanted.

```{r, warning=FALSE, message=FALSE}
world_map_temp_change(avg_country_temp_data)
```

--------

This chart shows the rise in occurences of natural disasters over time. Looking at the chart, we can see that the overall number of natural disasters has risen greatly. Starting in the early 1900's it is easy to confirm that the total number of natural disasters was below 100, even below 25 natural disasters a year. In 1975 we see the number of "All Natural Disasters" cross 100 in a steep upwards trend that continues all the way to the present where in 2012 we see over 400 natural disasters a year. This trend is highly correlated with the information in the plot below showing change in temperature correlating to change in number of natural disasters.

```{r, warning=FALSE, message=FALSE}
#plot_two, still from temp file
trendline_nd_count_by_year(natural_disaster_data)
```

--------

This chart shows the relationship between the average land temperature and the count of each natural disaster for that year. Added to the plot is the color coding of the amount of damage caused by the natural disasters. Important information to draw from the plots is the obvious trend that with a rise in temperature comes a rise in the number of each disaster. Looking at the damage caused by each disaster separately allows us to see which disasters in particular have a greater impact on people with the number of occurences rising. The "All Natural Disasters" plot clearly shows that the rise in temperature brings a rise in natural disasters and that the more natural disasters that occur, the more damage is done.

```{r, warning=FALSE, message=FALSE}
scatter_nd_count_by_temp(natural_disaster_data)
```

